---
title: "Statistical Analysis of Functional Capacity and Quality of Life in Pediatric Patients with Marfan Syndrome"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

# Setup and loading of required R packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
packages <- c("readxl", "dplyr", "gridExtra")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

# Step 7. Import datasets clean in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("../data/data_clean.xlsx")
excluded_data <- read_excel("../data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 8. Descriptive statistics of included and excluded patients at baseline

```{r}
library(dplyr)
library(gridExtra)
library(grid)

# Fonction descriptive
describe <- function(x) {
  data.frame(
    Statistic = c("Mean", "SD", "Median", "Q1", "Q3"),
    Value = c(
      mean(x, na.rm = TRUE),
      sd(x, na.rm = TRUE),
      median(x, na.rm = TRUE),
      quantile(x, 0.25, na.rm = TRUE),
      quantile(x, 0.75, na.rm = TRUE)
    )
  )
}

vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before",
          "vat_before",
          "pedsql_totalself_before")

labels <- c(
  "Age (years)",
  "Weight (kg)",
  "Height (cm)",
  "BMI (kg/m²)",
  "6MWT distance (m)",
  "VAT (% predicted)",
  "PedsQL score (0–100)"
)

table_list <- list()

for(i in seq_along(vars)) {
  
  var <- vars[i]
  
  x <- included_data[[var]]
  y <- excluded_data[[var]]
  
  # Remove NA (important pour le test)
  x_clean <- x[!is.na(x)]
  y_clean <- y[!is.na(y)]
  
  # Mann-Whitney
  p_val <- round(wilcox.test(x_clean, y_clean)$p.value, 3)
  
  inc <- describe(x)
  exc <- describe(y)
  
  df <- data.frame(
    Variable = c(labels[i], rep("", 4)),
    Statistic = inc$Statistic,
    Included = round(inc$Value, 1),
    Excluded = round(exc$Value, 1),
    p_value = c(p_val, rep("", 4))
  )
  
  table_list[[i]] <- df
}

table_final <- do.call(rbind, table_list)

# ---- SEX ----
p_sexe <- round(fisher.test(table(
  c(included_data$sexe, excluded_data$sexe),
  c(rep("Included", nrow(included_data)),
    rep("Excluded", nrow(excluded_data)))
))$p.value, 3)

sexe_inc <- table(included_data$sexe)
sexe_exc <- table(excluded_data$sexe)

sex_df <- data.frame(
  Variable = c("Sex (M/F)", rep("", length(sexe_inc)-1)),
  Statistic = names(sexe_inc),
  Included = as.numeric(sexe_inc),
  Excluded = as.numeric(sexe_exc),
  p_value = c(p_sexe, rep("", length(sexe_inc)-1))
)

table_final <- rbind(table_final, sex_df)

table_final

library(gridExtra)
library(grid)

# Ajuster hauteur automatiquement
n_rows <- nrow(table_final)

png(
  "../figures/Table_2_Comparison_of_included_and_excluded_patients_at_baseline.png",
  width = 900,
  height = 2 + n_rows * 42,
  res = 150
)

# Table sans quadrillage
tg <- tableGrob(
  table_final,
  rows = NULL,
  theme = ttheme_minimal(  
    core = list(
      fg_params = list(fontsize = 9)
    ),
    colhead = list(
      fg_params = list(fontsize = 10, fontface = "bold")
    )
  )
)

grid.arrange(
  textGrob(
    "Table 2. Comparison of included and excluded patients at baseline",
    gp = gpar(fontsize = 10, fontface = "bold")
  ),
  tg,
  ncol = 1,
  heights = c(0.2, 5)  
)

dev.off()
```
Descriptive statistics were computed for included and excluded patients to assess potential selection bias. Continuous variables were summarized using mean ± standard deviation and median [IQR].

A summary table was constructed to compare baseline characteristics between groups, with results formatted for clarity. The table was displayed in the R Markdown document and exported as a PNG for reporting.


# Step 9. Comparison of 6MWT Before and After Intervention.

Due to the small sample size and the potential non-normal distribution of the data, a non-parametric paired Wilcoxon signed-rank test was used to compare 6-minute walk distance before and after the intervention.

A paired comparison of 6-minute walk distance before and after the intervention was performed in included patients.


```{r wilcoxon-test}

# Paired Wilcoxon test
wilcox_test <- wilcox.test(
  included_data$`6minwt_distance_traveled_m_after`,
  included_data$`6minwt_distance_traveled_m_before`,
  paired = TRUE,
  exact = TRUE
)
wilcox_test

# Export Wilcoxon test results to a text file
sink("../results/wilcoxon_test_results.txt")
cat("Paired Wilcoxon Test (6MWT before vs after)\n\n")
print(wilcox_test)
sink()

```

# Step 10. Visualization of the 6 dimensions of the PedsQL self-questionnaire after 6 months 

## Boxplots of PedsQL scores before and after the intervention

```{r pedsql-plot}
library(dplyr)
library(tidyr)
library(ggplot2)

# Data PedsQL before/after
pedsql_data <- included_data %>%
  select(
    pedsql_totalself_before, pedsql_totalself_after,
    pedsql_physiqueself_before, pedsql_physiqueself_after,
    pedsql_emotionself_before, pedsql_emotionself_after,
    pedsql_relationself_before, pedsql_relationself_after,
    pedsql_ecoleself_before, pedsql_ecoleself_after,
    pedsql_psycosocialself_before, pedsql_psycosocialself_after
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Score") %>%
  mutate(
    Time = ifelse(grepl("before", Variable), "Before", "After"),
    Time = factor(Time, levels = c("Before", "After")),
    Dimension = case_when(
      grepl("total", Variable) ~ "Total Score",
      grepl("physique", Variable) ~ "Physical",
      grepl("emotion", Variable) ~ "Emotional",
      grepl("relation", Variable) ~ "Social",
      grepl("ecole", Variable) ~ "School",
      grepl("psycosocial", Variable) ~ "Psychosocial"
    )
  )

# Medians calculation for each dimension and time point
medians <- pedsql_data %>%
  group_by(Dimension, Time) %>%
  summarise(median = median(Score, na.rm = TRUE), .groups = "drop")

# Plot
p_pedsql <- ggplot(pedsql_data, aes(x = Time, y = Score, fill = Dimension)) +
  
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  
  # Medians lines each dimension
  geom_line(
    data = medians,
    aes(x = Time, y = median, group = Dimension, color = Dimension),
    size = 1
  ) +
  
  # Medians points
  
  geom_point(
    data = medians,
    aes(x = Time, y = median, color = Dimension),
    size = 2
  ) +
  
  facet_wrap(~ Dimension, nrow = 2) +
  
  xlab("") +
  ylab("Score (0–100)") +
  ggtitle("Figure 4. Distribution of PedsQL Scores Before and After Intervention") +
  
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"   
  )
p_pedsql

# Export
ggsave(
  "../figures/Figure_4_PedsQL_Before_After.png",
  plot = p_pedsql,
  width = 8,
  height = 6
)
```

# Step 11. Physical performance correlation – quality of life

The association between functional capacity and quality of life was assessed using Spearman’s rank correlation coefficient, due to the non-normal distribution of the data.

The analysis was performed between 6-minute walk distance and PedsQL score.

```{r}
# Spearman correlation
cor_test <- cor.test(
  included_data$`6minwt_distance_traveled_m_after`,
  included_data$pedsql_totalself_after,
  method = "spearman"
)

cor_test

# Export Spearman correlation results
sink("../results/spearman_correlation_results.txt")
cat("Spearman correlation (6MWT vs PedsQL)\n\n")
print(cor_test)
sink()

# Viewing
library(ggplot2)

# Create plot
p_corr <- ggplot(included_data, aes(
  x = `6minwt_distance_traveled_m_after`,
  y = pedsql_totalself_after
)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE) +
  xlab("6-Minute Walk Distance (m)") +
  ylab("PedsQL Score (0–100)") +
  ggtitle("Figure 5. Correlation Between Quality of Life and Functional Capacity measured at 6mwt ") +
  theme_minimal()

p_corr

# Export PDF
ggsave(
  "../figures/Figure_5_Correlation_PedsQL_6MWT.png",
  plot = p_corr,
  width = 8,
  height = 6
)
```



