---
title: "Data analysis project of the Marfan Power study as part of the teaching unit Movement Analysis"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Following preprocessing in Python, cleaned datasets were exported to R for statistical analysis.

Two datasets were used: included patients for the main analysis and excluded patients to assess potential bias.

All statistical analyses were performed in R.

# Step 8. Import datasets in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("data/data_clean.xlsx")
excluded_data <- read_excel("data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 9. Descriptive statistics of included and excluded patients

```{r table-comparison}

# Load packages
library(gridExtra)
library(grid)

# ---- 1. FUNCTIONS ----

# Mean ± SD
mean_sd <- function(x) {
  paste0(
    round(mean(x, na.rm = TRUE), 1), " ± ",
    round(sd(x, na.rm = TRUE), 1)
  )
}

# Median [IQR]
median_iqr <- function(x) {
  paste0(
    round(median(x, na.rm = TRUE), 1), " [",
    round(quantile(x, 0.25, na.rm = TRUE), 1), "–",
    round(quantile(x, 0.75, na.rm = TRUE), 1), "]"
  )
}

# ---- 2. VARIABLES ----

vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before",
          "pedsql_totalself_before")

# ---- 3. BUILD CLEAN TABLE ----

table_final <- data.frame(
  Variable = vars,
  Included = paste0(
    sapply(included_data[vars], mean_sd), "\n",
    sapply(included_data[vars], median_iqr)
  ),
  Excluded = paste0(
    sapply(excluded_data[vars], mean_sd), "\n",
    sapply(excluded_data[vars], median_iqr)
  )
)

# Remove row names
rownames(table_final) <- NULL

# Rename variables (clean labels)
table_final$Variable <- c(
  "Age (years)",
  "Weight (kg)",
  "Height (cm)",
  "BMI (kg/m²)",
  "6MWT distance (m)",
  "PedsQL score"
)

# Display table in RMarkdown
table_final

# ---- 4. EXPORT TO PDF ----

n_rows <- nrow(table_final)

pdf("figures/Table_2_Comparison_Included_Excluded.pdf",
    width = 8,
    height = 1 + n_rows * 0.5)

grid.arrange(
  textGrob(
    "Table 2. Comparison of included and excluded patients",
    gp = gpar(fontsize = 12, fontface = "bold")
  ),
  tableGrob(table_final),
  ncol = 1
)

dev.off()
```







