---
title: "Data analysis project of the Marfan Power study as part of the teaching unit Movement Analysis"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Following preprocessing in Python, cleaned datasets were exported to R for statistical analysis.

Two datasets were used: included patients for the main analysis and excluded patients to assess potential bias.

All statistical analyses were performed in R.

# Step 8. Import datasets in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("data/data_clean.xlsx")
excluded_data <- read_excel("data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 9. Descriptive statistics of included and excluded patients

```{r table-comparison}

# Load packages
library(gridExtra)
library(grid)

# Functions

# Mean ± SD
mean_sd <- function(x) {
  paste0(
    round(mean(x, na.rm = TRUE), 1), " ± ",
    round(sd(x, na.rm = TRUE), 1)
  )
}

# Median [IQR]
median_iqr <- function(x) {
  paste0(
    round(median(x, na.rm = TRUE), 1), " [",
    round(quantile(x, 0.25, na.rm = TRUE), 1), "–",
    round(quantile(x, 0.75, na.rm = TRUE), 1), "]"
  )
}

# Variables

vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before",
          "pedsql_totalself_before")

# Build clean table

table_final <- data.frame(
  Variable = vars,
  Included = paste0(
    sapply(included_data[vars], mean_sd), "\n",
    sapply(included_data[vars], median_iqr)
  ),
  Excluded = paste0(
    sapply(excluded_data[vars], mean_sd), "\n",
    sapply(excluded_data[vars], median_iqr)
  )
)

# Remove row names
rownames(table_final) <- NULL

# Rename variables 
table_final$Variable <- c(
  "Age (years)",
  "Weight (kg)",
  "Height (cm)",
  "BMI (kg/m²)",
  "6MWT distance (m)",
  "PedsQL score"
)

# Display table in RMarkdown
table_final

# Export to PDF

pdf("figures/Table_2_Comparison_Included_Excluded.pdf", width = 8, height = 7)

grid.arrange(
  textGrob(
    "Table 2. Comparison of included and excluded patients",
    gp = gpar(fontsize = 12, fontface = "bold")
  ),
  tableGrob(table_final, rows = NULL),
  ncol = 1,
  heights = c(0.3, 5) 
)
dev.off()

```

Descriptive statistics were computed for included and excluded patients to assess potential selection bias. Continuous variables were summarized using mean ± standard deviation and median [IQR].

A summary table was constructed to compare baseline characteristics between groups, with results formatted for clarity. The table was displayed in the R Markdown document and exported as a PDF for reporting.

## Interpretation of Baseline Characteristics (Table 2)

Baseline characteristics were compared between included and excluded patients to assess potential selection bias.

Overall, the two groups appeared comparable across most variables. Age was slightly higher in excluded patients (13.8 ± 2.8 years) compared to included patients (12.4 ± 4.0 years), although the distributions overlapped substantially. Similarly, anthropometric measures such as weight, height, and BMI were broadly similar between groups, with no clinically meaningful differences observed.

Baseline functional capacity, assessed by the 6-minute walk test (6MWT), was comparable between included and excluded patients (577.4 ± 55.4 m vs. 597.4 ± 24.6 m), suggesting no major difference in physical performance at baseline.

Quality of life scores (PedsQL) also showed overlapping distributions, although variability appeared greater in excluded patients.

Overall, these findings suggest that excluded patients were not markedly different from included patients at baseline, indicating a limited risk of major selection bias.


# Step 10. Comparison of 6MWT Before and After Intervention.

A paired comparison of 6-minute walk distance before and after the intervention was performed in included patients.

H0: There is no difference in 6-minute walk distance before and after the intervention.

H1: There is a difference in 6-minute walk distance before and after the intervention.



```{r wilcoxon-test}

# Paired Wilcoxon test
wilcox_test <- wilcox.test(
  included_data$`6minwt_distance_traveled_m_before`,
  included_data$`6minwt_distance_traveled_m_after`,
  paired = TRUE
)

wilcox_test
```
## Interpretation of Wilcoxon test

A statistically significant difference was observed between 6-minute walk distance before and after the intervention (Wilcoxon signed-rank test, V = 40, p = 0.016).

This finding suggests that the intervention was associated with a significant change in functional capacity.

