---
title: "Data analysis project of the Marfan Power study as part of the teaching unit Movement Analysis"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Following preprocessing in Python, cleaned datasets were exported to R for statistical analysis.

Two datasets were used: included patients for the main analysis and excluded patients to assess potential bias.

All statistical analyses were performed in R.

# Step 8. Import datasets in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("data/data_clean.xlsx")
excluded_data <- read_excel("data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 9. Descriptive statistics of included and excluded patients

```{r descriptive-stats}

# Function to compute descriptive statistics
describe_variable <- function(x) {
  c(
    Mean = round(mean(x, na.rm = TRUE), 1),
    SD = round(sd(x, na.rm = TRUE), 1),
    Median = round(median(x, na.rm = TRUE), 1),
    Q1 = round(quantile(x, 0.25, na.rm = TRUE), 1),
    Q3 = round(quantile(x, 0.75, na.rm = TRUE), 1)
  )
}

# Variables to compare (baseline)
vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before",
          "pedsql_totalself_before")

# Compute statistics
included_stats <- as.data.frame(t(sapply(included_data[vars], describe_variable)))
excluded_stats <- as.data.frame(t(sapply(excluded_data[vars], describe_variable)))

# Add group column
included_stats$Group <- "Included"
excluded_stats$Group <- "Excluded"

# Combine tables
final_stats <- rbind(included_stats, excluded_stats)

# Display final table
final_stats

# Create PDF
pdf("figures/Table_2_descriptive_stats_included_vs_excluded.pdf", width = 8, height = 4)

# Convert table to graphic
gridExtra::grid.table(final_stats)

# Close PDF
dev.off()
```






