---
title: "Data analysis project of the Marfan Power study as part of the teaching unit Movement Analysis"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Following preprocessing in Python, cleaned datasets were exported to R for statistical analysis.

Two datasets were used: included patients for the main analysis and excluded patients to assess potential bias.

All statistical analyses were performed in R.

# Step 8. Import datasets in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("data/data_clean.xlsx")
excluded_data <- read_excel("data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 9. Descriptive statistics of included and excluded patients

```{r import-data}
# Function to format Mean ± SD
mean_sd <- function(x) {
  paste0(
    round(mean(x, na.rm=TRUE),1), " ± ",
    round(sd(x, na.rm=TRUE),1)
  )
}

# Function to format Median [IQR]
median_iqr <- function(x) {
  paste0(
    round(median(x, na.rm=TRUE),1), " [",
    round(quantile(x,0.25,na.rm=TRUE),1), "-",
    round(quantile(x,0.75,na.rm=TRUE),1), "]"
  )
}

# Variables
vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before",
          "pedsql_totalself_before")

# Build table
table_final <- data.frame(
  Variable = vars,
  Included_MeanSD = sapply(included_data[vars], mean_sd),
  Excluded_MeanSD = sapply(excluded_data[vars], mean_sd),
  Included_MedianIQR = sapply(included_data[vars], median_iqr),
  Excluded_MedianIQR = sapply(excluded_data[vars], median_iqr)
)

table_final

library(gridExtra)
library(grid)

pdf("figures/Table_2_Comparison_Included_Excluded.pdf", width=10, height=5)

grid.arrange(
  textGrob("Table 2. Comparison of included and excluded patients",
           gp=gpar(fontsize=12, fontface="bold")),
  tableGrob(table_final),
  ncol=1
)

dev.off()
```







