---
title: "Data analysis project of the Marfan Power study as part of the teaching unit Movement Analysis, part (2/2)"
author: "Chloé Laignel-Granier"
date: "2026-04-14"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Following preprocessing in Python, cleaned datasets were exported to R for statistical analysis.

Two datasets were used: included patients for the main analysis and excluded patients to assess potential bias.

All statistical analyses were performed in R.

# Step 8. Import datasets in R

```{r import-data}
# Load package
library(readxl)

# Import datasets
included_data <- read_excel("data/data_clean.xlsx")
excluded_data <- read_excel("data/data_excluded.xlsx")

# Check structure
str(included_data)
str(excluded_data)
```
The cleaned datasets were imported into R for statistical analysis. Separate datasets were used for included patients, forming the main analytical sample, and excluded patients, allowing the evaluation of potential selection bias.

# Step 9. Descriptive statistics of included and excluded patients at baseline

```{r table-comparison}

# Load packages
library(gridExtra)
library(grid)

# Functions

# Mean ± SD
mean_sd <- function(x) {
  paste0(
    round(mean(x, na.rm = TRUE), 1), " ± ",
    round(sd(x, na.rm = TRUE), 1)
  )
}

# Median [IQR]
median_iqr <- function(x) {
  paste0(
    round(median(x, na.rm = TRUE), 1), " [",
    round(quantile(x, 0.25, na.rm = TRUE), 1), "–",
    round(quantile(x, 0.75, na.rm = TRUE), 1), "]"
  )
}

# Variables

vars <- c("age", "weight_kg", "height_cm", "bmi_kg_cm2",
          "6minwt_distance_traveled_m_before","vat_before",
          "pedsql_totalself_before")

# p-values

p_values <- sapply(vars, function(var) {
  
  x <- included_data[[var]]
  y <- excluded_data[[var]]
  
  # Remove NA
  x <- x[!is.na(x)]
  y <- y[!is.na(y)]
  
  # Quantitative → Mann-Whitney
  test <- wilcox.test(x, y)
  
  return(round(test$p.value, 3))
})

# Build clean table

table_final <- data.frame(
  Variable = vars,
  Included = paste0(
    sapply(included_data[vars], mean_sd), "\n",
    sapply(included_data[vars], median_iqr)
  ),
  Excluded = paste0(
    sapply(excluded_data[vars], mean_sd), "\n",
    sapply(excluded_data[vars], median_iqr)
  ),
  p_value = p_values
)

# Remove row names
rownames(table_final) <- NULL

# Rename variables 
table_final$Variable <- c(
  "Age (years)",
  "Weight (kg)",
  "Height (cm)",
  "BMI (kg/m²)",
  "6MWT distance (m)",
  "VAT (% predicted)",
  "PedsQL score (0-100)"
)

# Fisher test for sex 

fisher_test <- fisher.test(
  table(
    sexe = c(included_data$sexe, excluded_data$sexe),
    group = c(
      rep("Included", nrow(included_data)),
      rep("Excluded", nrow(excluded_data))
    )
  )
)

p_sexe <- round(fisher_test$p.value, 3)

# Counts
sexe_included <- table(included_data$sexe)
sexe_excluded <- table(excluded_data$sexe)

# Format (ex: M 12 / F 8)
sexe_included_txt <- paste(names(sexe_included), sexe_included, collapse = " / ")
sexe_excluded_txt <- paste(names(sexe_excluded), sexe_excluded, collapse = " / ")

# Add row to table
table_final <- rbind(
  table_final,
  data.frame(
    Variable = "Sex (M/F), n",
    Included = sexe_included_txt,
    Excluded = sexe_excluded_txt,
    p_value = p_sexe,
    stringsAsFactors = FALSE
  )
)

# Display table in RMarkdown
table_final

# Export to PDF

pdf("figures/Table_2_Comparison_Included_Excluded at baseline.pdf", width = 8, height = 7)

grid.arrange(
  textGrob(
    "Table 2. Comparison of included and excluded patients at baseline",
    gp = gpar(fontsize = 12, fontface = "bold")
  ),
  tableGrob(table_final, rows = NULL),
  ncol = 1,
  heights = c(0.3, 5) 
)
dev.off()
```

Descriptive statistics were computed for included and excluded patients to assess potential selection bias. Continuous variables were summarized using mean ± standard deviation and median [IQR].

A summary table was constructed to compare baseline characteristics between groups, with results formatted for clarity. The table was displayed in the R Markdown document and exported as a PDF for reporting.

## Interpretation of Baseline Characteristics (Table 2)

Baseline characteristics were compared between included and excluded patients to assess potential selection bias.

Overall, the two groups appeared comparable across most variables. Age was slightly higher in excluded patients (13.8 ± 2.8 years) compared to included patients (12.4 ± 4.0 years), although the distributions overlapped substantially. Similarly, anthropometric measures such as weight, height, and BMI were broadly similar between groups, with no clinically meaningful differences observed.

Baseline functional capacity, assessed by the 6-minute walk test (6MWT), was comparable between included and excluded patients (577.4 ± 55.4 m vs. 597.4 ± 24.6 m), suggesting no major difference in physical performance at baseline.

Quality of life scores (PedsQL) also showed overlapping distributions between groups, although greater variability was observed among excluded patients.

Sex distribution was similar between included (F 6 / M 14) and excluded patients (F 5 / M 3), with no statistically significant difference (Fisher’s exact test, p = 0.200).

Overall, these findings suggest that excluded patients were not markedly different from included patients at baseline, indicating a limited risk of major selection bias.


# Step 10. Comparison of 6MWT Before and After Intervention.

Due to the small sample size and the potential non-normal distribution of the data, a non-parametric paired Wilcoxon signed-rank test was used to compare 6-minute walk distance before and after the intervention.

A paired comparison of 6-minute walk distance before and after the intervention was performed in included patients.

H0: There is no difference in 6-minute walk distance before and after the intervention.

H1: There is a difference in 6-minute walk distance before and after the intervention.



```{r wilcoxon-test}

# Paired Wilcoxon test
wilcox_test <- wilcox.test(
  included_data$`6minwt_distance_traveled_m_before`,
  included_data$`6minwt_distance_traveled_m_after`,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)
wilcox_test

# Viewing individual changes at a 6-minute walking distance

library(ggplot2)

# Create plot
p <- ggplot(included_data, aes(x = delta_6mwt)) +
  geom_histogram(bins = 10, fill = "lightgrey", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  xlab("Change in 6-Minute Walk Distance (m)") +
  ylab("Number of patients") +
  ggtitle("Figure 3. Histogram of Individual Changes in the 6-Minute Walk Test") +
  theme_minimal()

# Display
p

# Save as PDF
ggsave(
  filename = "figures/Figure_3_Histogram_of_Individual_Changes_in_the_6-Minute_Walk_Test.pdf",
  plot = p,
  width = 6,
  height = 4
)
```

## Interpretation of Wilcoxon test

A paired Wilcoxon signed-rank test revealed a statistically significant difference in 6-minute walk distance before and after the intervention (V = 40, p = 0.016).

The median paired difference indicated an increase in 6MWT distance following the intervention. The 95% confidence interval for the median difference did not include zero, supporting the statistical significance of this change.

The distribution of individual changes in 6-minute walk distance is shown in Figure 3.

Most patients exhibited a positive change in 6MWT distance, indicating an overall improvement following the intervention. Only a small number of patients showed a decrease in performance.

The distribution appears slightly right-skewed, with some patients demonstrating large improvements, while a few showed minimal or negative changes. The dashed red line at zero highlights the threshold between improvement and decline.

Overall, this graphical representation supports the results of the Wilcoxon test, confirming a general increase in functional capacity after the intervention.

# Step 11. Physical performance correlation – quality of life

The association between functional capacity and quality of life was assessed using Spearman’s rank correlation coefficient, due to the non-normal distribution of the data.

The analysis was performed between 6-minute walk distance and PedsQL score.

```{r}
# Spearman correlation
cor_test <- cor.test(
  included_data$`6minwt_distance_traveled_m_after`,
  included_data$pedsql_totalself_after,
  method = "spearman"
)

cor_test

# Viewing
library(ggplot2)

# Create plot
p_corr <- ggplot(included_data, aes(
  x = `6minwt_distance_traveled_m_after`,
  y = pedsql_totalself_after
)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  xlab("6-Minute Walk Distance (m)") +
  ylab("PedsQL Score (0–100)") +
  ggtitle("Figure 4. Correlation Between Functional Capacity and Quality of Life") +
  theme_minimal()

p_corr

# Export PDF
ggsave(
  "figures/Figure_4_Correlation_6MWT_PedsQL.pdf",
  plot = p_corr,
  width = 6,
  height = 4
)
```
## Interpretation of Spearman correlation

A Spearman correlation analysis was performed to assess the relationship between functional capacity and quality of life.

A weak positive correlation was observed between 6-minute walk distance and PedsQL score (Spearman’s rho = 0.19). However, this association was not statistically significant (p = 0.430).

The scatter plot shows a slight upward trend, suggesting that higher functional capacity may be associated with better quality of life. Nevertheless, the wide dispersion of the data points indicates substantial variability between individuals.

Overall, these results do not support a significant association between functional capacity and quality of life in this sample.

This lack of statistical significance may be explained by the limited sample size and inter-individual variability.











```


